---
title: "Introduction to R"
subtitle: "Import and Clean Data in R"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, hygge, ninjutsu]
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r, include = FALSE}
# This is the recommended set up for flipbooks
# you might think about setting cache to TRUE as you gain practice --- building flipbooks from scratch can be time consuming
knitr::opts_chunk$set(fig.width = 6, message = FALSE, warning = FALSE, comment = "", cache = F)

library(flipbookr)
library(tidyverse)
library(here)
library(haven)
```

## Welcome!

This tutorial will demonstrate how to get started in R and R Studio

To follow along with the workshop in R, you will need to have **already installed** R and R Studio

---

## How to follow along and participate

* Open R Studio

* Install the `tidyverse`, `rmarkdown`, `haven`, and `here` packages (if you haven't already)

    + Use the `Tools` menu and choose the `Install packages...` menu choice 

    + Type the names of the packages, `tidyverse, rmarkdown`, in the dialog box that opens and click Install

* Open the workshop R Markdown file and click on the **tiny** green arrow on the top right corner of the gray shaded area (this is a code chunk!) that contains this: 

```{r eval = FALSE}
library(package = "here")
library(package = "tidyverse")
library(package = "haven")
```

* The arrow will run all the code in that "chunk" of code

* Throughout the workshop you will be able to run the code from the demonstration by clicking the arrow on the top right in the corresponding code chunk

* In between code chunks, you can type as you would in a Word document if you'd like to take notes

* For the exercises you may have to finish the code yourself and then click the arrow to get the anticipated result

---

## Why are we doing this?

* R is open source and free 

* R can do everything that SPSS and SAS can do 

* R is especially great for data visualization 

* The R community is supportive and active 

---

## The data set

* We will be using the **Global Attitudes & Trends** data set from the **Pew Internet & American Life** website 

    + If needed, download the data and documentation from this website: <a href = "https://www.pewresearch.org/global/datasets/">https://www.pewresearch.org/global/datasets/</a>
    
---

## Importing data into R

* Before running this code, make sure the participant file and the data file are saved in the same folder on your computer

```{r}
# use here so that R looks in the same folder you are in
here()

# use read_sav function to read the spss file 
globalData2019 <- read_spss(file = "Pew Research Center Global Attitudes Spring 2019 Dataset WEB.sav")
```

---

## Examining the data set 

* After importing the data you should see **globalData2019** in your **Environment** tab 

* Next to the name you gave to the data, **globalData2019**, you will see `r nrow(globalData2019)` obs. of `r ncol(globalData2019)` variables 

    + This means there are `r nrow(globalData2019)` observations, which in this case are the survey participants 
    
    + It also means there are `r ncol(globalData2019)` variables 
    
* The codebook for these data describe the `r ncol(globalData2019)` variables

* Click one time on the name of the data **globalData2019**, and it will pop-up in a data viewer tab in the **Console** pane

---

## Choosing some variables and observations to examine 

* Making a smaller dataset is often useful if there is a subset of people you are interested in, or even just to make the data easier to easily view

* For this workshop, we will choose:

    + 7 variables: SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY
    
    + Observations from Nigeria (country 21)

---

`r chunk_reveal("small_data", widths = c(1,0), title = "## Selecting variables and filtering observations")`

```{r small_data, include = FALSE}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) 

```

---

## Examining the smaller data set

```{r}
summary(object = globalData2019clean)
```

---

## Planning data cleaning 

* All of these variables except `AGE` should be categorical, which is a **factor** data type in R 

* The way the variables are shown in the `summary()` indicates R thinks they are **numeric** data type 

* Most of these variables also have categories or responses like **Don't know** or **Refused** that are probably not useful for analysis 

* Data cleaning should include: 

    + Changing the categorical variables into the factor type 
    
    + Recoding the responses like **Don't know** or **Refused** to be `NA` in R
    
        + `NA` stands for **Not Available** in R and is how a missing value is recorded 

---

`r chunk_reveal("data_types", widths = c(1,0), title = "## Changing data types 1")`

```{r data_types, include = FALSE}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female'))
```

---

## Checking your work 1

```{r}
summary(object = globalData2019clean)
```

---

`r chunk_reveal("data_types2", widths = c(1,0), title = "## Changing data types 2")`

```{r data_types2, include = FALSE}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                       '1' = 'Men',
                                       '2' = 'Women',
                                       '3' = 'Same/both equally',
                                       '8' = NA_character_,
                                       '9' = NA_character_))
```

---

## Checking your work 2

```{r}
summary(object = globalData2019clean)
```

---

`r chunk_reveal("data_types3", widths = c(1,0), title = "## Changing data types 3")`

```{r data_types3, include = FALSE}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_))
```

---

## Checking your work 3

```{r}
summary(object = globalData2019clean)
```

---

## Recoding data: You Try It!

* Look in the codebook to find the variable for **ECON_SIT** (note: it will be lower case in the codebook)

* Add recoding of **ECON_SIT** to the code below and run the code 

* Run the `summary()` code in the next code chunk to check your work

```{r include = FALSE, echo = FALSE}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) 
```


```{r echo = FALSE, include = FALSE}
# check your work
summary(object = globalData2019clean)
```

---

`r chunk_reveal("you_try_key", widths = c(1,0), title = "## Answer to You Try It!")`

```{r you_try_key, include = FALSE}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) %>% 
  mutate(ECON_SIT = recode_factor(ECON_SIT,
                                  '1' = 'Very good',
                                  '2' = 'Somewhat good',
                                  '3' = 'Somewhat bad',
                                  '4' = 'Very bad',
                                  '8' = NA_character_,
                                  '9' = NA_character_))
```

---

## Check your work: You Try It!

* If the code ran properly, you should expect to see the four categories listed under the `ECON_SIT` variable name

* If you do not see this, go back to your code and look for missing commas and other punctuation problems 

```{r}
summary(object = globalData2019clean)
```

---

## Try Some More! 

* Recode the last two categorical variables, `BELIEVE_GOD` and `GLOBAL_COMMUNITY` to have appropriate category labels and `NA` 

* Use the codebook to find the labels you need 

* Add to the existing code: 

```{r include = FALSE}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) %>% 
  mutate(ECON_SIT = recode_factor(ECON_SIT,
                                  '1' = 'Very good',
                                  '2' = 'Somewhat good',
                                  '3' = 'Somewhat bad',
                                  '4' = 'Very bad',
                                  '8' = NA_character_,
                                  '9' = NA_character_))
```

---

## Check your work: Try Some More!


* If the code ran properly, you should expect to see the categories listed under the variable names

* If you do not see this, go back to your code and look for missing commas and other punctuation problems 


---

`r chunk_reveal("try_more_key", widths = c(1,0), title = "## Answer to Try Some More!")`

```{r try_more_key, include = FALSE}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) %>% 
  mutate(ECON_SIT = recode_factor(ECON_SIT,
                                  '1' = 'Very good',
                                  '2' = 'Somewhat good',
                                  '3' = 'Somewhat bad',
                                  '4' = 'Very bad',
                                  '8' = NA_character_,
                                  '9' = NA_character_)) %>% 
  mutate(BELIEVE_GOD = recode_factor(BELIEVE_GOD,
                                     '1' = 'It is not necessary to believe in God in order to be moral and have good values',
                                     '2' = 'It is necessary to believe in God in order to be moral and have good values',
                                     '8' = NA_character_,
                                     '9' = NA_character_)) %>% 
  mutate(GLOBAL_COMMUNITY = recode_factor(GLOBAL_COMMUNITY,
                                          '1' = 'Part of a global community that works together to solve problems',
                                          '2' = 'Independent nations that compete with other countries and pursue their own interests',
                                          '3' = 'Both',
                                          '4' = 'Neither',
                                          '8' = NA_character_,
                                          '9' = NA_character_))
```

---

## Check your work: Try Some More!

```{r}
summary(object = globalData2019clean)
```

---

## Final recoding for age

```{r}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>%
  mutate(AGE = na_if(AGE, 98)) %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) %>% 
  mutate(ECON_SIT = recode_factor(ECON_SIT,
                                  '1' = 'Very good',
                                  '2' = 'Somewhat good',
                                  '3' = 'Somewhat bad',
                                  '4' = 'Very bad',
                                  '8' = NA_character_,
                                  '9' = NA_character_)) %>% 
  mutate(BELIEVE_GOD = recode_factor(BELIEVE_GOD,
                                     '1' = 'It is not necessary to believe in God in order to be moral and have good values',
                                     '2' = 'It is necessary to believe in God in order to be moral and have good values',
                                     '8' = NA_character_,
                                     '9' = NA_character_)) %>% 
  mutate(GLOBAL_COMMUNITY = recode_factor(GLOBAL_COMMUNITY,
                                          '1' = 'Part of a global community that works together to solve problems',
                                          '2' = 'Independent nations that compete with other countries and pursue their own interests',
                                          '3' = 'Both',
                                          '4' = 'Neither',
                                          '8' = NA_character_,
                                          '9' = NA_character_)) 

```

---

## The end

* Email me at [harrisj@wustl.edu](mailto:harrisj@wustl.edu)

* Connect with me on Twitter [@jenineharris](https://twitter.com/jenineharris)




<!-- adjust font size in this css code chunk, currently 80 -->

```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1.5; font-size: 80%}

@media print {
  .has-continuation {
    display: block;
  }
}

code.r.hljs.remark-code{
  position: relative;
  overflow-x: hidden;
}


code.r.hljs.remark-code:hover{
  overflow-x:visible;
  width: 500px;
  border-style: solid;
}
```




